---
title: "Decomposing pseudo-bulk DE into transcriptional and compositional
  components with scCompoundDE"
author:
  - name: "Subhadip Jana"
    email: "subhadipjana1409@gmail.com"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Decomposing pseudo-bulk DE with scCompoundDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse  = TRUE,
    comment   = "#>",
    fig.width = 7,
    fig.height = 5,
    message   = TRUE,
    warning   = FALSE
)
```

# Introduction

## The hidden confound in pseudo-bulk DE

Multi-donor single-cell RNA-seq studies routinely use **pseudo-bulk
differential expression** to find disease or treatment-associated genes.
The standard approach aggregates counts per donor and fits a linear model
across donors. This is statistically sound — but it contains a critical
biological blind spot.

Consider T cells in a lupus dataset. Disease donors have mostly exhausted
T cells (high PDCD1, LAG3, TIGIT). Healthy donors have mostly naive T
cells (high IL7R, CCR7, SELL). A pseudo-bulk DE model will declare PDCD1
and IL7R as the top DE genes — and both would have small p-values and
large fold-changes. But:

> **Are these genes DE because T cells changed their transcriptome
> (transcriptional change)?**
>
> **Or because the relative abundance of exhausted vs naive T cells
> shifted between conditions (compositional change)?**

Standard tools cannot answer this question. They treat the pseudo-bulk
as a homogeneous entity. `scCompoundDE` decomposes every DE gene into
these two orthogonal components.

## The TC_ratio score

For each gene, `compoundDE` returns a `TC_ratio` in \[0, 1\]:

| TC_ratio | Interpretation |
|---|---|
| ≈ 1.0 | Transcriptional — cells intrinsically changed expression |
| ≈ 0.0 | Compositional — proportion of subtypes shifted |
| ≈ 0.5 | Mixed — both mechanisms contribute |

Genes with low TC_ratio should be treated with caution: they may be
DE in a standard analysis but reflect **composition** rather than
**biology**.

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("scCompoundDE")
```

# Simulated example: purely transcriptional signal

We first demonstrate the tool on a dataset where the DE signal is
deliberately **transcriptional**: the same proportions of TypeA and
TypeB cells appear in both conditions, but TypeA cells intrinsically
upregulate genes 1–10 in treatment.

```{r sim_transcriptional}
library(scCompoundDE)
library(SingleCellExperiment)

set.seed(42)
n_genes  <- 150
n_donors <- 6

counts <- matrix(rpois(n_genes * 240L, 8L),
                 nrow = n_genes, ncol = 240L)
rownames(counts) <- paste0("Gene", seq_len(n_genes))
colnames(counts) <- paste0("Cell", seq_len(240L))

# First 10 genes × 4 in TypeA treatment cells only
# TypeA cells: cols 61–120 in treatment
counts[seq_len(10), 61:120] <- counts[seq_len(10), 61:120] * 5L

sce <- SingleCellExperiment(assays = list(counts = counts))
sce$donor     <- rep(rep(paste0("D", seq_len(n_donors)), each = 10L), 4L)
sce$cell_type <- "T_cell"
sce$subtype   <- rep(rep(c("TypeA","TypeB"),
                         each = n_donors * 10L), 2L)
sce$condition <- rep(c("ctrl","treat"), each = 120L)

# Check the design
table(sce$subtype, sce$condition)
```

## Running compoundDE

```{r run_compoundDE}
result <- compoundDE(
    sce,
    broad_type   = "T_cell",
    subtype_col  = "subtype",
    broad_col    = "cell_type",
    donor        = "donor",
    condition    = "condition",
    min_cells    = 3L,
    min_subtypes = 2L,
    min_donors   = 2L
)
result
```

## Inspecting results

```{r inspect_results}
dt <- as.data.frame(deTable(result))
dt_sorted <- dt[order(dt$adj.P.Val), ]
head(dt_sorted[, c("gene","logFC","adj.P.Val","TC_ratio","source")], 15)
```

The injected genes (Gene1–Gene10) should have high TC_ratio values,
confirming they were detected as transcriptionally driven.

```{r tc_of_injected}
injected <- dt[dt$gene %in% paste0("Gene", seq_len(10)), ]
cat("Median TC_ratio of injected genes:",
    round(median(injected$TC_ratio, na.rm = TRUE), 3), "\n")
cat("All injected genes classified as transcriptional:",
    all(injected$source == "transcriptional", na.rm = TRUE), "\n")
```

## Visualisations

```{r plot_decomp, fig.cap="Scatter of transcriptional vs compositional z-scores. Blue = transcriptional, red = compositional, orange = mixed. Injected genes appear in the top-right (high T, low C)."}
plotDecomposition(result, fdr_thresh = 0.05, top_n = 10)
```

```{r plot_tc, fig.cap="Distribution of TC_ratio scores. Vertical lines show classification thresholds at 0.2 and 0.8."}
plotTCRatio(result, fdr_thresh = 0.05)
```

```{r plot_prop, fig.cap="Subtype proportions per condition. Balanced design — equal TypeA and TypeB in both conditions."}
plotProportion(result)
```

# Simulated example: purely compositional signal

Now we demonstrate the tool on a dataset where the **apparent** DE is
caused entirely by a subtype proportion shift. Gene1 is a marker of
TypeB cells. In control, most cells are TypeA (low Gene1). In
treatment, most cells become TypeB (high Gene1). **A standard DE tool
would call Gene1 as significantly upregulated — but it is a compositional
artefact.**

```{r sim_compositional}
set.seed(99)
n_genes <- 150

# Unequal proportions: ctrl = 75% TypeA, treat = 75% TypeB
n_a_ctrl  <- 72L;  n_b_ctrl  <- 24L
n_a_treat <- 24L;  n_b_treat <- 72L
n_cells   <- n_a_ctrl + n_b_ctrl + n_a_treat + n_b_treat

counts2 <- matrix(rpois(n_genes * n_cells, 5L),
                  nrow = n_genes, ncol = n_cells)
rownames(counts2) <- paste0("Gene", seq_len(n_genes))
colnames(counts2) <- paste0("Cell", seq_len(n_cells))

# Gene1 is a marker of TypeB only — no transcriptional change
typeB_idx <- c(
    seq(n_a_ctrl + 1L, n_a_ctrl + n_b_ctrl),
    seq(n_a_ctrl + n_b_ctrl + n_a_treat + 1L, n_cells)
)
counts2[1L, typeB_idx] <- counts2[1L, typeB_idx] * 8L

sce2 <- SingleCellExperiment(assays = list(counts = counts2))
sce2$donor <- c(
    rep(paste0("D", 1:6), each = 12L),
    rep(paste0("D", 1:6), each = 4L),
    rep(paste0("D", 1:6), each = 4L),
    rep(paste0("D", 1:6), each = 12L)
)
sce2$cell_type <- "T_cell"
sce2$subtype   <- c(rep("TypeA", n_a_ctrl), rep("TypeB", n_b_ctrl),
                    rep("TypeA", n_a_treat), rep("TypeB", n_b_treat))
sce2$condition <- c(rep("ctrl",  n_a_ctrl + n_b_ctrl),
                    rep("treat", n_a_treat + n_b_treat))

table(sce2$subtype, sce2$condition)
```

```{r run_compositional}
result2 <- compoundDE(
    sce2,
    broad_type   = "T_cell",
    subtype_col  = "subtype",
    broad_col    = "cell_type",
    donor        = "donor",
    condition    = "condition",
    min_cells    = 3L,
    min_subtypes = 2L,
    min_donors   = 2L
)
result2
```

```{r compositional_check}
dt2 <- as.data.frame(deTable(result2))
gene1 <- dt2[dt2$gene == "Gene1", c("logFC","adj.P.Val","TC_ratio","source")]
cat("Gene1 (compositional marker):\n")
print(gene1)
```

Gene1 will have a significant p-value (broad DE is strong) but a low
TC_ratio, correctly identifying it as a compositional artefact.

```{r plot_comp_decomp, fig.cap="With compositional signal, Gene1 sits at high C_score / low T_score."}
plotDecomposition(result2, fdr_thresh = 0.05, top_n = 5)
```

```{r plot_comp_prop, fig.cap="Clear proportion shift: TypeB increases dramatically in treatment."}
plotProportion(result2)
```

# Filtering by source

After running `compoundDE`, use `filterGenesBySource` to extract gene
lists by classification:

```{r filter_genes}
# Genes with transcriptionally-driven DE (real biology)
transcriptional_genes <- filterGenesBySource(result,
    source     = "transcriptional",
    fdr_thresh = 0.05)
cat("Transcriptional genes (FDR < 0.05):", nrow(transcriptional_genes), "\n")
head(transcriptional_genes[, c("gene","logFC","adj.P.Val","TC_ratio")])

# Genes that are artefactual (compositional)
compositional_genes <- filterGenesBySource(result,
    source     = "compositional",
    fdr_thresh = 1)    # include all, regardless of significance
cat("\nCompositional genes (any FDR):", nrow(compositional_genes), "\n")
```

# Recommended workflow

```{r workflow, eval=FALSE}
# Step 1: QC with scBatchQC
library(scBatchQC)
sce <- batchAwareQCMetrics(sce, batch = "batch")
sce_clean <- sce[, !sce$scBatchQC_outlier]

# Step 2: Fast DE with scFastDE (or any pseudo-bulk tool)
library(scFastDE)
de_result <- fastDE(sce_clean,
                    donor       = "donor",
                    cell_type   = "cell_type",
                    condition   = "condition",
                    target_type = "T_cell")

# Step 3: Decompose with scCompoundDE
cde_result <- compoundDE(
    sce_clean,
    broad_type  = "T_cell",
    subtype_col = "cell_subtype",
    broad_col   = "cell_type",
    donor       = "donor",
    condition   = "condition"
)

# Step 4: Retain only transcriptionally-driven genes
real_biology <- filterGenesBySource(cde_result,
    source = "transcriptional", fdr_thresh = 0.05)
```

# Session information

```{r session}
sessionInfo()
```
